An Experimental Study of Sorting and Branch Prediction

### 2006-01-20 docs/branch_paper/paper.tex
done initial pictures

\usepackage{graphicx}

@@

@@

Although we know the approximate number of comparisons that each
algorithm performs, and important question is whether there is any
difference of the predictability of the branches in each algorithm.
Figure \ref{} shows the number of mispredictions per key for both the
algorithms using random data, and varying data set sizes, measured
using the Simplescalar processor simulator.

@@

Figure \ref{} shows the number of branch mispredictions per key for
bubble- and shakersort.  Nearly one in four comparisons in bubblesort
are mispredicted. As the data sets gets large, there are nearly 10000
branch mispredictions per key key. Shakersort executes a lot fewer
branches, and so has fewer mispredictions, but the rate of
misprediction is similar.

@@

\begin{figure}
\includegraphics{plots/two_point_counter_bubblesort.eps}
\caption{Behaviour of branches in bubblesort compared to the ``sweep'' number}
\label{Predictability-of-branches-in-bubblesort-compared-to-sweep-number}
\end{figure}

@@

The main apparent weakness of shellsort on modern architectures is its
poor locality of reference in memory access.  By comparing items which
are separated by $h$ keys, there is little reuse of items in the same
cache line. One can easily imagine cases where shellsort would cause
very large numbers of cache misses.  In the worst case, each key would
move a very large number of places in each insertion sort pass,
possibly causing a cache miss for every place that it moves.  To
further investigate this effect, we measured the number of first and
second-level cache misses caused by our implementation of shellsort
(see figure \ref{}). In fact, the number of mispredictions caused by
our shellsort is significant, but not much greater than other sorting
algorithms such as heapsort and merge sort (see sections \ref{} and
\ref{}). The main reason is that using Gonnet's increments on average
keys move only 0.9744 places on each pass, so very large numbers of
references in diverse parts of the array are not needed.

\section{Heapsort}
Heapsort is an in-place sort which works in $O(NlogN)$ time in the
average and worst case. It begins by creating a structure known as a
\n{heap}, a binary tree in which a parent is always larger its child,
and the largest key is at the top of the heap. It then repeatedly
removes the largest key from the top and places it in its final
position in the array, fixing the heap as it goes.

@@

\subsection{Multi-Quicksort}
Generally, quicksort has good locality of reference. It divides the
same region of the array into increasingly small partitions.  In the
case that the array to be sorted fits in a particular level of the
cache, then quicksort has excellent cache properties at that
level. Once the array is larger than the cache, however, the number of
misses increases dramatically. On the early passes, which divide the
array into large sections, there is likely to be little reuse of data
cache entries between passes. Given that there is little reuse between
passes, it would be better to reduce the number of passes by dividing
the array into a larger number of partitions on each pass. LaMarca
uses a technique similar to that used in multi-mergesort, in that a
\n{multi-way partition} is done at the start, moving all keys into
cache sized containers which can be sorted individually.

>>>>

\usepackage{graphicx}
\usepackage{subfigure}

@@

\newcommand{\subscale}{0.50}
\newcommand{\mainscale}{1}

@@

Although we know the approximate number of comparisons that each
algorithm performs, and important question is whether there is any
difference of the predictability of the branches in each algorithm.
Figure \ref{Predictability-of-insertion-and-selection-sorts} shows the number of mispredictions per key for both the
algorithms using random data, and varying data set sizes, measured
using the Simplescalar processor simulator.

\begin{figure}[h]
\includegraphics[scale=\mainscale]{plots/bpred_misses_insert_select.eps}
\caption{TODO: Predictability of insertion and selection sorts}
\label{Predictability-of-insertion-and-selection-sorts}
\end{figure}

@@

Figure \ref{Predictability-of-bubblesort-and-shakersort} shows the number of branch mispredictions per key for
bubble- and shakersort.  Nearly one in four comparisons in bubblesort
are mispredicted. As the data sets gets large, there are nearly 10000
branch mispredictions per key key. Shakersort executes a lot fewer
branches, and so has fewer mispredictions, but the rate of
misprediction is similar.

\begin{figure}[h]
\includegraphics[]{plots/bpred_misses_bubble_shaker.eps}
\caption{TODO: Predictability of bubble and shaker sorts}
\label{Predictability-of-bubblesort-and-shakersort}
\end{figure}

@@

\begin{figure}[h]
\includegraphics{plots/bpred_bubble_progress.eps}
\caption{Behaviour of branches in bubblesort compared to the ``sweep'' number}
\label{Predictability-of-branches-in-bubblesort-compared-to-sweep-number}
\end{figure}

@@

The main apparent weakness of shellsort on modern architectures is its
poor locality of reference in memory access.  By comparing items which
are separated by $h$ keys, there is little reuse of items in the same
cache line. One can easily imagine cases where shellsort would cause
very large numbers of cache misses.  In the worst case, each key would
move a very large number of places in each insertion sort pass,
possibly causing a cache miss for every place that it moves.  To
further investigate this effect, we measured the number of first and
second-level cache misses caused by our implementation of shellsort
(see figure \ref{Cache-misses-in-shellsort}). In fact, the number of mispredictions caused by
our shellsort is significant, but not much greater than other sorting
algorithms such as heapsort and merge sort (see sections \ref{} and
\ref{}). The main reason is that using Gonnet's increments on average
keys move only 0.9744 places on each pass, so very large numbers of
references in diverse parts of the array are not needed.

\begin{figure}[h]
\subfigure[Caption1]{\includegraphics[scale=\subscale]{plots/cache_misses_level_2_shell.eps}}
\subfigure[Caption2]{\includegraphics[scale=\subscale]{plots/cache_misses_level_2_shell.eps}}
\caption{TODO: Predictability of insertion and selection sorts}
\label{Cache-misses-in-shellsort}
\end{figure}

\section{Heapsort}
Heapsort is an in-place sort which works in $O(NlogN)$ time in the
average and worst case. It begins by creating a structure known as a
heap, a binary tree in which a parent is always larger its child,
and the largest key is at the top of the heap. It then repeatedly
removes the largest key from the top and places it in its final
position in the array, fixing the heap as it goes.

@@

\subsection{Multi-Quicksort}
Generally, quicksort has good locality of reference. It divides the
same region of the array into increasingly small partitions.  In the
case that the array to be sorted fits in a particular level of the
cache, then quicksort has excellent cache properties at that
level. Once the array is larger than the cache, however, the number of
misses increases dramatically. On the early passes, which divide the
array into large sections, there is likely to be little reuse of data
cache entries between passes. Given that there is little reuse between
passes, it would be better to reduce the number of passes by dividing
the array into a larger number of partitions on each pass. LaMarca
uses a technique similar to that used in multi-mergesort, in that a
multi-way partition is done at the start, moving all keys into
cache sized containers which can be sorted individually.
